A geometric approach to linear Chabauty

### 2023-12-08 MSK.tex
Add defense files

\newcommand{\den}{\mathop{\mathgroup\symoperators den}\nolimits}
\newcommand{\ggd}{\mathop{\mathgroup\symoperators ggd}\nolimits}
\newcommand{\des}{\text{d.e.s.d.a.\ }}
%\newcommand{\opg}{\text{d.e.s.d.a.\ }}
\newcommand{\blokje}{\hfill $\Box$\\}
%Om je bewijs of je antwoord af te sluiten.
\newcommand{\header}[1]{\vspace{0.5cm}\framebox[\linewidth]{\textsc{Exercise #1}}\\}
\newcommand{\deel}[1]{\textbf{#1)}\ }
\newcommand{\macht}[1]{\mathcal{P}(#1)}
\newcommand{\entier}[1]{\left\lfloor #1 \right\rfloor}
\newcommand{\A}{\mathbb{A}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\renewcommand{\G}{\mathbb{G}}
\renewcommand{\C}{\mathbb{C}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\Fcal}{\mathcal{F}}
\newcommand{\Lcal}{\mathcal{L}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\Pcal}{\mathcal{P}}
\newcommand{\NP}{\mathcal{NP}}
\newcommand{\NPC}{\mathcal{NPC}}
\newcommand{\F}{\mathbb{F}}
\renewcommand{\angle}[1]{\hspace{-2pt}\left\langle #1 \right\rangle}
\newcommand*\diff{\mathop{}\!\mathrm{d}}
\newcommand{\rad}{\text{rad}}
\newcommand{\x}{x}
\newcommand{\m}{\mathfrak{m}}
\newcommand{\tensor}{\otimes}
\newcommand{\Ring}{\textbf{Rings}}
\newcommand{\Sets}{\textbf{Sets}}

\DeclareMathOperator{\mon}{mon}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Gal}{Gal}
\DeclareMathOperator{\End}{End}
\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\trdeg}{trdeg}
\DeclareMathOperator{\rk}{rk}
\DeclareMathOperator{\Fun}{Fun}
\DeclareMathOperator{\im}{im}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\sm}{sm}
\DeclareMathOperator{\pr}{pr}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\MaxSpec}{MaxSpec}
\DeclareMathOperator{\Spf}{Spf}
\DeclareMathOperator{\sep}{sep}
\DeclareMathOperator{\Ann}{Ann}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\Mat}{Mat}
\let\S\undefined
\DeclareMathOperator{\S}{S}
\let\div\undefined
\DeclareMathOperator{\div}{div}

@@

\title{A geometric approach to linear and quadratic Chabauty}
\author{Pim Spelier}
\date{November 2015}

@@

{\LARGE\bf 
A geometric approach to linear Chabauty
}

\vspace{10em}

@@

\vspace{10em}

{\large\bf
\begin{tabular}{ll}
Thesis supervisor: & prof. dr. S.J. Edixhoven\\
\end{tabular}
}

\vfill

@@

\end{center}
\thispagestyle{empty}
\newpage
\listoftodos
\tableofcontents
\newpage

%-----------------------------------------------------------------------------------------------------------------------
\todo{Remove overfull h-box warnings}
\section{Introduction}
\label{section:intro}
In mathematics a fundamental problem is solving polynomial equations over the rationals, dating back to Diophantus. An important special case in algebraic geometry is that of curves. The behavior of the rational points of curves depends enormously on the \textit{genus} $g$ of the curve $C$, a numerical invariant of a curve. For $g = 0$, there are either no or infinitely many solutions, and their behavior is well understood. For $g = 1$, we get either no point or an elliptic curve, whose set of rational points forms an often infinite group.

In this thesis, we look only at the case $g > 1$. It turns out that there are always only finitely many rational points; this result was originally conjectured by Mordell in 1922 and finally proven by Faltings in 1983.

Before it was proven, one of the major partial results was a theorem from Chabauty in 1941, phrased in terms of the rank $r$ of the group of rational points on the Jacobian $J$ of the curve, also called the Mordell-Weil rank, which is finite by the Mordell-Weil theorem. Chabauty proved, using $p$-adic methods, that if $r$ is strictly smaller than the genus $g$, then $C$ has only finitely many rational points. Loosely said, this proof and all Chabauty related theorems, rely on choosing a prime $p$ for which $C$ has good reduction, and intersecting $J(\Z)$ with $C(\Z_p)$ inside the bigger Lie group $J(\Z_p)$; by properties of $\Z_p$, the subgroup $J(\Z)$, up to torsion generated by $r$ elements, lies within a Lie subgroup of dimension at most $r$, and $C(\Z_p)$ is a $1$-dimensional $p$-adic manifold, and hence their intersection can be proven to be discrete and -- as $J(\Z_p)$ is compact -- finite.\todo{Ik moet volgens mij uberhaupt nadenken wat over $\Q,\Q_p,\Z,\Z_p$ moet}

This was later made into an effective argument by Coleman in 1985. This was done by finding explicitly, as a power series, a differential form $\omega$ on $J_{\Z_p}$ whose Coleman integral vanishes on $J(\Z)$. Pulling $\omega$ back to $C_{\Z_p}$ and looking at the fibre of reduction to a single $\F_p$-point $P$, one can in certain cases give an upper bound for the number of points in $C(\Z)$ reducing to $P$ based only on finite-precision calculations of $\omega$. Most well known is the generic upper bound $|C(\Q)| \leq |C(\F_p)| + 2g-2$ under the conditions $r < g, p > 2g$, as explained in the introductory article \cite{poonen12}.

This Coleman-Chabauty method has been greatly generalised by Kim to so called non-abelian Chabauty in \todo{Ref} and \todo{Ref}. He interprets working in the Jacobian as dealing with the abelianised fundamental group of the curve $C$, and works with larger, non-abelian quotients of the fundamental group. Then the simplest case of non-abelian Chabauty, quadratic Chabauty was developed by Kim, Balakrishnan, Besser Dogran and M\"uller in the series of articles \todo{Ref}, and further extended by Balakrishnan, Dogra, M\"uller, Tuitman and Vonk; in 2017 they famously calculated  all rational points of the ``cursed curve'', the modular curve $X_s^+(13)$ \cite{cursedcurve}. They use an endomorphism of the Jacobian, and do $p$-adic analysis on $p$-adic local heights to make quadratic Chabauty explicit. Their methods are strong enough to prove finiteness of $C(\Z)$ under the condition of $r < g + \rho - 1$, where $\rho$ is the N\'eron-Severi rank of the Jacobian.

Very recently, Edixhoven and Lido made available a preprint of their article ``Geometric quadratic Chabauty'' \cite{edixhoven20}. Their goal in this article is to make a more geometric approach to effective quadratic Chabauty, by working in a pullback $T$ of the Poincar\'e torsor of the Jacobian. They also abandon the Coleman method for Chabauty; instead they parametrise the map $T(\Z) \to T(\Z_p)$ with power series, and pull back equations for the curve inside $T(\Z_p)$ along this map. Again, a necessary condition is $r < g + \rho - 1$.

The purpose of this thesis is to make the work by Edixhoven and Lido more accessible by applying geometric Chabauty to the linear case, i.e. when working with $J(\Z)$ and $J(\Z_p)$ and assuming $r < g$. In this thesis, we go into more detail of what is happening throughout the process, including both theoretical and practical components. This new method of applying Chabauty leads to, when working with single digit $p$-adic precision, a conditional upper bound $|C(\Q)| \leq |C(\F_p)|$, as obtained in Proposition~\ref{prop:linchabauty}. This upper bound compares favorably with classical Coleman-Chabauty, but unlike classical Coleman-Chabauty, we cannot say in general when the conditions of these propositions hold. We develop heuristics that suggest that generally trying multiple primes $p$ of good reduction will yield good results; these heuristics say that the conditional upper bound will hold for a set of primes of density $1$, and if $r < g-1$ can even often (in a set of primes of density at least $e^{-1}$) be improved to prove there are at most $|C(\Q)|$ rational points. The methods used to obtain Proposition~\ref{prop:linchabauty} are also amenable to higher precision calculations; indeed, in the case $r = g-1$ this may be necessary to calculate $|C(\Z)|$ exactly.\todo{Over hebben met Bas}

We also give a hyperelliptic example with $r = 1, g = 2$, where the bound from Proposition~\ref{prop:linchabauty} indeed holds, and we are able to compute $C(\Z)$.

\subsection{Overview}
We first present the context in which we will perform Chabauty. Let $p>2$ be a prime, and $C$ be a scheme over $\Z$, proper, flat, regular with $C$ smooth over $\Z_{(p)}$ and $C_\Q$ of dimension $1$ and geometrically connected, of genus $g \geq 1$ and with the N\'eron model $J$ of the Jacobiano of $C_\Q$ having Mordell-Weil rank $r < g$. Assume we have a $\Q$-point $b$ in $C$, or equivalently a $\Z$-point. We view $C$ as a subscheme of $J$, using the map $Q \mapsto Q - b$ on points. Let $P \in C(\F_p)$ be a point such that $t := P - b \in J(\F_p)$ lies in the image of $J(\Z)$.

\begin{defn}
Let $S$ be a scheme, $T \to U$ a morphism of schemes and $x : T \to S$ a $T$-point. We define $S(U)_x$ as the morphisms from $U$ to $S$ that, after precomposing with $T \to U$, give $x$.
\end{defn}
\begin{exmp}
If we have a projective variety $X$ over $\Z$, then $X(\Z)$ is in bijection with $X(\Q)$. The natural map $X(\Z) \to X(\F_p)$ reduces a point modulo $p$ and for $x \in X(\F_p)$, the set $X(\Z)_x$ consists of the $\Z$-points reducing to $x$.
\end{exmp}

Geometric Chabauty works by finding an upper bound for the cardinality of $C(\Z)_P$. For this, we use the following diagram.
\[
\begin{tikzcd}
C(\Z)_P \arrow[d] \arrow[r, "a"] & J(\Z)_t \arrow[d] \\
C(\Z_p)_P \arrow[r]              & J(\Z_p)_t        
\end{tikzcd}
\]
where the two vertical maps are inclusions, and the two horizontal maps are subtraction of $b$. We will in fact compute upper bounds for the larger set $C(\Z_p)_P \cap \overline{J(\Z)_t}$, where $\overline{J(\Z)_t}$ is the closure of $J(\Z)_t$ in $J(\Z_p)_t$.

In Section~\ref{section:smoothzppoints}, we treat the structure of $C(\Z_p)_P$ and $J(\Z_p)_t$; we will show that, after choosing parameters, they are canonically in bijection with respectively $\Z_p$ and $\Z_p^g$. We also discuss the resulting map $\Z_p \to \Z_p^g$.

In Section~\ref{section:kappa} we further look at the map $J(\Z)_t \to J(\Z_p)_t$. This is a translation of the group morphism $J(\Z)_0 \to J(\Z_p)_0$ between kernels of reduction; it turns out that for our choice of $p$, the subgroup $J(\Z)_0$ is free of rank $r$. We study the properties of the resulting map $\Z^r \to \Z_p^g$, using the theory of formal groups to determine the group structure on $\Z_p^g$ induced by the bijection $J(\Z_p)_t \to \Z_p^g$.

In Section~\ref{section:intersection}, we put all of this information together, culminating in several methods to possibly compute upper bounds on $\left|C(\Z_p)_P \cap \overline{J(\Z)_t}\right|$ with finite precision calculations. We also perform a heuristic analysis to show that a simple calculation of linear algebra modulo $p$ is very likely to result in an upper bound of at most $|C(\F_p)|$ for $|C(\Z)|$, and for $r < g-1$ even in an upper bound of $|C(\Z)|$.

The methods developed in the first four sections are not always guaranteed to prove finiteness of $C(\Z)_P$. In Section~\ref{section:remarks} we treat several possible complications and improvements, referring to other work on Chabauty for further reading.

Next, in Section~\ref{section:explicit} we focus on how to perform the calculations happening in the Jacobian. Here, we use Makdisi's approach of representing a divisor by a subspace of a large Riemann-Roch space. We partially follow the article \cite{mascot18} by Mascot, and also give an original algorithm for explicitly computing the map $C(\Z_p)_P \to J(\Z_p)_t$ in Makdisi's representation.

We end our discussion of linear geometric Chabauty with an explicit example in Section~\ref{section:example}, a genus $2$ curve whose Jacobian has Mordell-Weil rank $1$. We use both Magma and Pari/GP to do our calculations, and end up with a complete list of all rational points of the curve.

Finally, we provide an introduction to geometric quadratic Chabauty in Section~\ref{section:quadratic}. Here, we aim to introduce all the different objects relevant to geometric quadratic Chabauty; we explain why we have to work over $\Z$ instead of over $\Q$ to make geometric quadratic Chabauty make sense; and we explain some of the problems and solutions that working over $\Z$ gives rise to.

\subsection{Preliminaries}
We expect the readers of this thesis to be familiar with algebraic geometry, e.g. master students with a specialisation in algebraic geometry. The introduction to geometric quadratic Chabauty in Section~\ref{section:quadratic} contains some terms typically not treated in a master; while knowing them will help put the material in context, this section is meant to be understandible even if that is not the case.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
% Say something about smooth schemes over Z_p, and their Z_p points
\section{Points on a smooth scheme over \texorpdfstring{$\Z_p$}{Z\_p}}
\label{section:smoothzppoints}
Let $X/\Z_p$ be a smooth scheme of relative dimension $d$, and let $x \in X(\F_p)$ be a point. Then $X(\Z_p)_x$ is in bijection with $\Z_p^d$. This bijection is given by choosing parameters at $x$; evaluating at $X(\Z_p)_x$ gives a bijection with $(p\Z_p)^d$, and then we divide by $p$. For putting up a nice framework to work in, we start with blowing up $X$ at $x$.

% Blow up X at x, and look at the part where p generates the maximal ideal
Assume, by looking at a neighborhood of $x$, that $X = \Spec A$ is affine and that $p,t_1,\dots,t_d$ generate the maximal ideal of $O_{X,x}$ with $t_1,\dots,t_d$ elements of $O_X(X) = A$, also called \textit{parameters} at $x$. By shrinking $X$ even more, we may assume as $X$ is smooth that $t = (t_1,\dots,t_d): X \to \A^d_{\Z_p}$ is \'etale. Now consider the blowup $\widetilde{X}_x \to X$ of $X$ at $x$, and let $\widetilde{X}_x^p$ be the part where $p$ generates the inverse image of the maximal ideal of $O_{X,x}$. Equivalently, that is the part where $t_1,\dots,t_d$ are multiples of $p$, so informally it consists of the points that reduce to $x$ modulo $p$.

There is an explicit description of the map $\widetilde{X}_x^p \to X$; as $t$ is \'etale, the ideal of $X$ defining $x$ is the pullback along $t$ of the ideal of $\A^d_{\Z_p}$ defining the origin $a$ over $\F_p$. That means that the blowup $\widetilde{X}_x \to X$ is the pullback of the blowup $\widetilde{\A}^d_{\Z_p,a} \to \A_{\Z_p}^d$. Then the part $\widetilde{X}_x^p$ is the pullback of the corresponding part of $\widetilde{\A}^d_{\Z_p,a}$, i.e. $\Spec \Z_p[\widetilde{x_1},\dots,\widetilde{x_d}] = \Spec \Z_p[x_1/p,\dots,x_d/p]$ with the morphism $\Z_p[x_1,\dots,x_d] \to \Z_p[\widetilde{x_1},\dots,\widetilde{x_d}]$ given by $x_i \mapsto p\widetilde{x_i}$. That implies that $\widetilde{X}_x^p$ is $\Spec A[t_1/p,\dots,t_d/p]$, with the map $\widetilde{X}_x^p \to X$ given by the inclusion $A \to \Spec A[t_1/p,\dots,t_d/p]$ (remember that the $t_i$ are elements of $O_X(X) = A$).

>>>>

\newcommand{\den}{\mathop{\mathgroup\symoperators den}\nolimits}
\newcommand{\ggd}{\mathop{\mathgroup\symoperators ggd}\nolimits}
\newcommand{\des}{\text{d.e.s.d.a.\ }}
%\newcommand{\opg}{\text{d.e.s.d.a.\ }}
\newcommand{\blokje}{\hfill $\Box$\\}
%Om je bewijs of je antwoord af te sluiten.
\newcommand{\header}[1]{\vspace{0.5cm}\framebox[\linewidth]{\textsc{Exercise #1}}\\}
\newcommand{\deel}[1]{\textbf{#1)}\ }
\newcommand{\macht}[1]{\mathcal{P}(#1)}
\newcommand{\entier}[1]{\left\lfloor #1 \right\rfloor}
\newcommand{\A}{\mathbb{A}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\renewcommand{\G}{\mathbb{G}}
\renewcommand{\C}{\mathbb{C}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\va}{\mathbf{a}}
\newcommand{\vx}{\mathbf{x}}
\newcommand{\vy}{\mathbf{y}}
\newcommand{\vz}{\mathbf{z}}
\newcommand{\Fcal}{\mathcal{F}}
\newcommand{\Lcal}{\mathcal{L}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\Pcal}{\mathcal{P}}
\newcommand{\NP}{\mathcal{NP}}
\newcommand{\NPC}{\mathcal{NPC}}
\newcommand{\F}{\mathbb{F}}
\renewcommand{\angle}[1]{\hspace{-2pt}\left\langle #1 \right\rangle}
\newcommand*\diff{\mathop{}\!\mathrm{d}}
\newcommand{\rad}{\text{rad}}
\newcommand{\x}{x}
\newcommand{\m}{\mathfrak{m}}
\newcommand{\tensor}{\otimes}
\newcommand{\Ring}{\textbf{Rings}}
\newcommand{\Sets}{\textbf{Sets}}

\DeclareMathOperator{\mon}{mon}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Gal}{Gal}
\DeclareMathOperator{\End}{End}
\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\trdeg}{trdeg}
\DeclareMathOperator{\rk}{rk}
\DeclareMathOperator{\Fun}{Fun}
\DeclareMathOperator{\im}{im}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\sm}{sm}
\DeclareMathOperator{\pr}{pr}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\MaxSpec}{MaxSpec}
\DeclareMathOperator{\Spf}{Spf}
\DeclareMathOperator{\sep}{sep}
\DeclareMathOperator{\Ann}{Ann}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\Mat}{Mat}
\DeclareMathOperator{\Jac}{Jac}
\let\S\undefined
\DeclareMathOperator{\S}{S}
\let\div\undefined
\DeclareMathOperator{\div}{div}

@@

\title{A geometric approach to linear Chabauty}
\author{Pim Spelier}
\date{November 2015}

@@

{\LARGE\bf 
A geometric approach to linear Chabauty\\
{\color{red} Over number fields}
}

\vspace{8em}

@@

\vspace{8em}

{\large\bf
\begin{tabular}{ll}
Thesis supervisor: & prof. dr. S.J. Edixhoven\\
\end{tabular}
}
\vfill

@@

\end{center}
\thispagestyle{empty}
\newpage
\tableofcontents
\newpage

%-----------------------------------------------------------------------------------------------------------------------
\section{Introduction}
\label{section:intro}
A fundamental problem in mathematics is solving polynomial equations over the rationals, dating back to Diophantus. An important special case in algebraic geometry is that of curves. The behavior of the rational points of curves depends enormously on the \textit{genus} $g$ of the curve, a numerical invariant. For $g = 0$, there are either no or infinitely many solutions, and their behavior is well understood. For $g = 1$, we get either no point or an elliptic curve, whose set of rational points forms an often infinite group.

In this thesis, we look only at the case of a curve $C$ with genus $g > 1$. It turns out that there are always only finitely many rational points. This result was originally conjectured by Mordell in 1922 and finally proven by Faltings in 1983 in \cite{faltings}.

Before Faltings's theorem was proven, one of the major partial results was a theorem from Chabauty in 1941, phrased in terms of the rank $r$ of the group of rational points on the Jacobian $J$ of the curve, also called the Mordell-Weil rank, which is finite by the Mordell-Weil theorem. Chabauty proved, using $p$-adic methods, that if $r$ is strictly smaller than the genus $g$, then $C$ has only finitely many rational points. Loosely said, this proof and all Chabauty related theorems, rely on choosing a prime $p$ for which $C$ has good reduction, and intersecting $J(\Q)$ with $C(\Q_p)$ inside the bigger Lie group $J(\Q_p)$. By properties of $\Q_p$, the subgroup $J(\Q)$, up to torsion generated by $r$ elements, lies within a Lie subgroup of dimension at most $r$, and $C(\Q_p)$ is a $1$-dimensional $p$-adic manifold, and hence this intersection can be proven to be discrete and -- as $J(\Q_p)$ is compact -- finite.

This was made into an effective argument by Coleman in 1985. This was done by finding explicitly, as a power series, a differential form $\omega$ on $J_{\Q_p}$ whose Coleman integral vanishes on $J(\Q)$. Pulling $\omega$ back to $C_{\Q_p}$ and looking at the fibre of reduction to a single $\F_p$-point $P$, one can in certain cases give an upper bound for the number of points in $C(\Q)$ reducing to $P$ based only on finite-precision calculations of $\omega$. Most well known is the generic upper bound $|C(\Q)| \leq |C(\F_p)| + 2g-2$ under the conditions $r < g, p > 2g$, as explained in the excellent introductory article \cite{poonen12}.

This Coleman-Chabauty method has been greatly generalised by Kim to so-called non-abelian Chabauty in \cite{kim1} and \cite{kim2}. He interprets working in the Jacobian as dealing with the abelianised fundamental group of the curve $C$ and works with larger, non-abelian quotients of the fundamental group. Then quadratic Chabauty, the simplest case of non-abelian Chabauty, was developed by Kim, Balakrishnan, Besser, Dogra and M\"uller in the series of articles \cite{qc1},\cite{qc2},\cite{qc3}, and further extended by Balakrishnan, Dogra, M\"uller, Tuitman and Vonk; in 2017 they famously calculated all rational points of the ``cursed curve'', the modular curve $X_s^+(13)$ \cite{cursedcurve}. They use an endomorphism of the Jacobian and do $p$-adic analysis on $p$-adic local heights to make quadratic Chabauty explicit. Their methods are strong enough to prove finiteness of $C(\Q)$ under the condition of $r < g + \rho - 1$, where $\rho$ is the N\'eron-Severi rank of the Jacobian, even if finding $C(\Q)$ can still be difficult in those cases.

Very recently, Edixhoven and Lido made available a preprint of their article ``Geometric quadratic Chabauty'' \cite{edixhoven20}. Their goal in this article is to create a more geometric approach to effective quadratic Chabauty, by working in a pullback $T$ of the Poincar\'e torsor of the Jacobian spread out over $\Z$. They also abandon the Coleman method for Chabauty; instead, they parametrise the map $T(\Z) \to T(\Z_p)$ with power series and pull back equations for the curve inside $T(\Z_p)$ along this map. Again, a necessary condition is $r < g + \rho - 1$.

The purpose of this thesis is to make the work by Edixhoven and Lido more accessible by applying geometric Chabauty to the linear case, i.e. when working with $J(\Q)$ and $J(\Q_p)$ and assuming $r < g$.\todo{Add number fields in the introduction} In this thesis, we go into more detail on what is happening throughout the process, including both theoretical and practical components. This new method of applying Chabauty leads to, when working with single-digit $p$-adic precision, a conditional upper bound $|C(\Q)| \leq |C(\F_p)|$, as obtained in Proposition~\ref{prop:linchabauty}. This upper bound compares favorably with classical Coleman-Chabauty, but unlike classical Coleman-Chabauty, we cannot say in general when the conditions of these propositions hold. We develop heuristics that suggest that generally, trying multiple primes $p$ of good reduction will yield good results; these heuristics say that the conditional upper bound will hold for a set of primes of density $1$, and if $r < g-1$ the upper bound can even often (in a set of primes of density at least $e^{-1}$) be improved to prove there are at most $|C(\Q)|$ rational points, i.e., finding $C(\Q)$ exactly. The methods used to obtain Proposition~\ref{prop:linchabauty} are also amenable to higher precision calculations; indeed, in the case $r = g-1$ one expects to need either higher precision calculations or the Mordell-Weil sieve to calculate $|C(\Q)|$ exactly.

We also give a hyperelliptic example with $r = 1, g = 2$, where the bound from Proposition~\ref{prop:linchabauty} indeed holds, and we are able to compute $C(\Q)$.

\subsection{Overview}
We first present the context in which we will perform Chabauty. Let $K$ be a number field of degree $d$, and let $v: K \to \Z$ be a finite place. Let $\pi$ denote a uniformiser, $q$ the cardinality of the residue field and $p> 2$ its caracteristic\todo{Check when $p,\pi$ is big enough for both Pierre Parent and logarithm}\todo{Decide if we can allow ramification}. Let $f$ denote the residue field degree $[\F_q : \F_p]$ and $e$ the ramification $v(p)$. We use $O$ to denote the maximal order of $K$, and $O_{(\pi)}$ and $O_{\pi}$ for respectively the localisation and the completion with respect to $\pi$. This valuation also gives rise to an embedding of $K$ into an extension $K_{\pi}$ of $\Q_p$ of degree $ef$, with ramification degree $e$ and residue field degree $f$; then $O_\pi$ is simply the ring of integers of $K_{\pi}$.

Let $C_{K}/K$ be a curve (i.e., a proper, smooth, geometrically connected variety of dimension $1$) of genus $g \geq 2$ whose $K$-points we will attempt to find. assume we have a scheme $C$ over $O_{(\pi)}$, proper and smooth with generic fibre $C_{K}$; then we immediately have $C_K(K) = C(K) = C(O_{(\pi)})$ by the valuative criterion of properness. Let $J$ be the relative Jacobian of $C$ over $O_{(\pi)}$, i.e. with fibres $\Jac C_K$ and $\Jac C_{\F_q}$. We assume from now on that the Mordell-Weil rank of $J$ is $r \leq d(g-1)$. Assume we have a $K$-point $b$ in $C$, or equivalently a $O_{(\pi)}$-point. We view $C$ as a subscheme of $J$, using the map $Q \mapsto Q - b$ on points. Let $P \in C(\F_q)$ be a point such that $t := P - b \in J(\F_q)$ lies in the image of $J(O_{(\pi)})$.

\begin{defn}
Let $S$ be a scheme, $T \to U$ a morphism of schemes and $x : T \to S$ a $T$-point. We define $S(U)_x$ as the morphisms from $U$ to $S$ that, after precomposing with $T \to U$, give $x$.
\end{defn}
\begin{exmp}
If we have a proper variety $X$ over $O_{(\pi)}$, then $X(O_{(\pi)})$ is naturally in bijection with $X(K)$. The natural map $X(O_{(\pi)}) \to X(\F_q)$ reduces a point modulo $\pi$ and for $x \in X(\F_q)$, the set $X(O_{(\pi)})_x$ consists of the residue disc of $O_{(\pi)}$-points reducing to $x$.
\end{exmp}

Geometric Chabauty works by finding an upper bound for the cardinality of $C(O_{(\pi)})_P$. For this, we use the following diagram.
\[
\begin{tikzcd}
C(O_{(\pi)})_P \arrow[d] \arrow[r] & J(O_{(\pi)})_t \arrow[d] \\
C(O_\pi)_P \arrow[r]              & J(O_\pi)_t        
\end{tikzcd}
\]
where the two vertical maps are inclusions, and the two horizontal maps are subtraction of $b$. We will in fact compute upper bounds for the larger set $C(O_\pi)_P \cap \overline{J(O_{(\pi)})_t}$, where $\overline{J(O_{(\pi)})_t}$ is the closure of $J(O_{(\pi)})_t$ in $J(O_\pi)_t$.

In Section~\ref{section:smoothzppoints}, we treat the structure of $C(O_\pi)_P$ and $J(O_\pi)_t$; we will show that, after choosing parameters, they are in bijection with respectively $O_\pi$ and $O_\pi^g$. We also discuss the resulting map $O_\pi \to O_\pi^g$.{\color{red} and, identifying $O_\pi$ with $\Z_p^{ef}$ as topological groups, the map $\Z_p^{ef} \to \Z_p^{efg}$. This latter map is not necessarily well behaved if there is ramification; choosing a $\Z_p$-basis of $O_{\pi}$ appropriately, we will instead look at the well behaved restriction $\Z_p^{f} \to \Z_p^{fg}$.}

In Section~\ref{section:kappa} we further look at the map $J(O_{(\pi)})_t \to J(O_\pi)_t$. This is a translation of the group morphism $J(O_{(\pi)})_0 \to J(O_\pi)_0$ between kernels of reduction; it turns out that for our choice of $p$, the subgroup $J(O_{(\pi)})_0$ is free of rank $r$. We study the properties of the resulting map $\Z^r \to O_\pi^g$, using the theory of formal groups to determine the group structure on $O_\pi^g$ induced by the bijection $J(O_\pi)_0 \to O_\pi^g$. {\color{red} Importantly, we identify $O_\pi$ with $\Z_p^{ef}$ as topological groups; we also look at the resulting map $\Z^r \to \Z_p^{efg}$. Again, to control the behavior of this map modulo $p$, we consider the map $\Z^r \to \Z_p^{fg}$.}

In Section~\ref{section:intersection}, we put all of this information together, culminating in several methods to possibly compute upper bounds on $\left|C(O_\pi)_P \cap \overline{J(O_{(\pi)})_t}\right|$ with finite precision calculations if $r \leq f(g-1)$. We also perform a heuristic analysis to show that a simple calculation of linear algebra modulo $p$ is very likely to result in an upper bound of at most $|C(\F_q)|$ for $|C(O_{(\pi)})|$, and for $r < f(g-1)$ \todo{check whether this is correct, it should hold.} even in an upper bound of $|C_K(K)|$. Hence given only a curve $C_K/K$, the best possible choice of a prime of $K$ is an inert prime

The methods developed in the first four sections are not always guaranteed to prove finiteness of $C(O_{(\pi)})_P$. In Section~\ref{section:remarks} we treat several possible complications and improvements, referring to other work on Chabauty for further reading.

Next, in Section~\ref{section:explicit} we focus on how to perform the calculations happening in the Jacobian. Here, we use Makdisi's approach of representing a divisor by a subspace of a large Riemann-Roch space. We partially follow the article \cite{mascot18} by Mascot, and also give an original algorithm for explicitly computing the map $C(O_{\pi})_P \to J(O_{\pi})_t$ in Makdisi's representation.

We end our discussion of linear geometric Chabauty with an explicit example in Section~\ref{section:example}, a genus $2$ curve over the rationals whose Jacobian has Mordell-Weil rank $1$. We use both Magma and Pari/GP to do our calculations, and end up with a complete list of all rational points of the curve.

Finally, we provide an introduction to geometric quadratic Chabauty in Section~\ref{section:quadratic} (working only over the rationals). Here, we aim to introduce all the different objects relevant to geometric quadratic Chabauty; we explain why we have to work over $\Z$ instead of over $\Q$ in order for geometric quadratic Chabauty to make sense; and we explain some of the problems and solutions that working over $\Z$ gives rise to.

\subsection{Preliminaries}
We expect the readers of this thesis to be familiar with algebraic geometry, e.g. master students with a specialisation in algebraic geometry. The introduction to geometric quadratic Chabauty in Section~\ref{section:quadratic} contains some terms typically not treated in a master programme; while knowing them will help to put the material in context, this section is meant to be understandable for those that are not familiar with the terminology as well.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
% Say something about smooth schemes over Z_p, and their Z_p points
\section{Points on a smooth scheme over \texorpdfstring{$\Z_p$}{Z\_p}}
\label{section:smoothzppoints}
{\color{red} Most of this section should still hold in the general number field case; replacing $p$ by $\pi$ when it is an element of $A$ (i.e., a section of $X$) and by $q$ when it is the cardinality of the residue field. Only the remark on it being a torsor of the tangent space merits a checking+rewrite. This has been done.}
Let $X/\Z_p$ be a smooth scheme of relative dimension $d$, and let $x \in X(\F_p)$ be a point. Then we will show in this section that $X(\Z_p)_x$ is naturally in bijection with $\Z_p^d$. This bijection is given by choosing parameters at $x$; evaluating at $X(\Z_p)_x$ gives a bijection with $(p\Z_p)^d$, and then we divide by $p$. For putting up a nice framework to work in, we start with blowing up $X$ at $x$.

% Blow up X at x, and look at the part where p generates the maximal ideal
Assume, by looking at a neighborhood of $x$, that $X = \Spec A$ is affine and that $p,t_1,\dots,t_d$ generate the maximal ideal of $\O_{X,x}$ with $t_1,\dots,t_d$ elements of $\O_X(X) = A$, also called \textit{parameters} at $x$. By shrinking $X$ even more, we may assume as $X$ is smooth that $t = (t_1,\dots,t_d): X \to \A^d_{\Z_p}$ is \'etale and the fibre of the origin over $\F_p$ consists of just $x$. Now consider the blowup $\widetilde{X}_x \to X$ of $X$ at $x$, and let $\widetilde{X}_x^p$ be the open subscheme where $p$ generates the inverse image of the maximal ideal of $\O_{X,x}$. Equivalently, that is the part where $t_1,\dots,t_d$ are multiples of $p$, so informally $\widetilde{X}_x^p$ consists of the points that reduce to $x$ modulo $p$.
